---
title: "Object Oriented Programming in R"
author: "Abel"
date: "18 Mar 2016"
output: html_document
---

# Introduction to object oriented programming in R, with special emphasis on the ExpressionSet class
Hector Corrada Bravo
based on slides developed by
Jim Bullard, Kasper Hansen and Margaret Taub

## Object Oriented Programming in R

```{r}
rm(list=ls(all.names = TRUE)); try(dev.off())
xx <- rnorm(1000)
class(xx)
plot(xx)
yy <- ecdf(xx)
class(yy)
plot(yy)
plot
plot.ecdf
plot.default
methods("plot")
getS3method("plot", "histogram")
```

What plot does, depends on the class of the x argument. It is a method. plot.ecdf is the ecdf method for plot.


## Constructing a new S3 Class
```{r}
jim <- list(height = 2.54 * 12 * 6/100, weight = 180/2.2, name = "James")
class(jim) <- "person2"
class(jim)
```
We have now made an object of class person. We now define a print method.

```{r}
print(jim)
print.person2 <- function(x, ...) {
	cat("name:", x$name, "\n")
	cat("height:", x$height, "meter s", "\n")
	cat("weight:", x$weight, "kilograms", "\n")
	}
print(jim)
```

Note the method/class has the "dot" naming convention of method.class.

## S3 classes are not robust

```{r}
fit <- lm(rnorm(100) ~ 1)
class(fit)
print(fit)
class(fit) <- "something"
# print(fit)
class(fit) <- "person2"
# print(fit)
```

In case print does not have a method for the class, it dispatches to the default method, print.default.
S3 does not have the concept of type checking { there is no way to formally define a class and ensure that the object conform to the definition.


## S3 classes and the help system

S3 classes are traditionally documented in the help page for the function that creates them. Example: lm.
Methods have a generic help page (often not very informative), sometimes with more specific help under ?method.class. Example:
plot.lm.


## Inheritance in S3

In S3, inheritance is achieved by the class attribute being a vector. A canonical example is
```{r}
fit <- glm(rpois(100, lambda = 1) ~
	1, family = "poisson")
class(fit)
methods("residuals")
methods("model.matrix")
```

If no method for the first is found, the second class is checked.


## Useful S3 Method Functions

- methods("print") and methods(class = "lm")
- getS3method("print","person") : Gets the appropriate method associated with a class, useful to see how a method is implemented.
	- Try: getS3method("residuals", "lm").

```{r}
methods("print")[1:10]
methods(class = "lm")
getS3method("residuals", "lm")
```

I Sometimes, methods are non-visible, because they are hidden in a namespace. Use getS3method or getAnywhere to get around this.

```{r}
# residuals.HoltWinters
# getS3method("residuals.HoltWinters")
getAnywhere("residuals.HoltWinters")
```

## Replacement Methods

I As we have already seen R has a somewhat strange type of function that allows us to modify objects in place.
I It is uncommon to define new replacement functions, however they are used quite frequently in day to day programming of R.
I Two examples are: names and colnames. Type \colnames" into the R  window and hit \tab", notice the function \colnames<-"?

```{r}
a <- matrix(1:16, nrow = 4, ncol = 4)
colnames(a) <- paste("V", 1:4, sep = ".")
colnames(a)
point <- list(x = 1, y = 2)
x.val <- function(x, value) { 		x$x <- value	}
"x.val<-" <- function(x, value) {	x$x <- value; return(x)	}


x.val(point,1)
# The 2nd argumetn does not do anything
print(point)
x.val(point) <- 10
print(point)
```

What does the first print statement print? What about the second?


## S4 classes, why?

- Although S3 classes can be quite useful and powerful and fast they do not facilitate the type of modularization and type safety that a true object oriented system intends.
- S4 classes are more a traditional object oriented system with type checking, multiple-dispatch, and inheritance.
- S4 is implemented in the methods package in base R.
- For thorough information on S4, read Chambers (1998)  \Programming with data" (also known as the green book) (first chapter available at http://www.omegahat.org/RSMethods/Intro.pdf) or Chambers (2008) \Software for Data Analysis: Programming with R".
- There are also several good, short, tutorials on the net.

## Defining an S4 class

```{r}
myRep <- representation(height = "numeric",	weight = "numeric", name = "character")
setClass("personS4", representation = myRep)
getClass("personS4")
jimS4 <- new("personS4")
jimS4
jimS4 <- new("personS4", height = 2.54 *12 * 6/100, weight = 180/2.2,	name = "James")
jimS4
jimS4@name
validObject(jimS4)
# jimS4@height <- "2"
jimS4@height <- 2

```

## Notes on the S4 class example

- It is rare for users to define their own S4 classes.
- The use of new to instantiate a new member of the class is not
always needed, often there are explicit constructor functions (see
later).
- The use of @ to access the class slots is heavily discouraged, instead
use accessor functions (see later).

## Defining the print method

For completion, we define the print method for personS4. For S4 classes, it is not print, but rather show.

```{r}
setMethod("show", signature("personS4"),
	function(object) {
	cat("name:", object@name,"\n")
	cat("height:", object@height, "meters", "\n")
	cat("weight:", object@weight, "kilograms", "\n")
	})
jimS4 # equivalent to / shorthand of show (jimS4)
getMethod("show", signature("personS4"))
```

## S4 Generics

In order to make a new generic we need to call the function setGeneric.

```{r}
setGeneric("BMI", function(object) standardGeneric("BMI"))
setMethod("BMI", "personS4", function(object) {
	object@weight/object@height^2
	})
BMI(jimS4)
```
